---
title: "Cox regression modeling of survival after chemotherapy for colon cancer"
author: "Hermela Shimelis"
date: today
format:
  html:
    code_folding: true
    css: custom.css
    toc: true
    toc_float:
      collapsed: TRUE
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
csl: ieee.csl
self-contained: true
embed-resources: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72 
---

## I. Introduction

The Cox Proportional Hazards (PH) model is one of the most widely used
regression models used for survival analysis, also known as
time-to-event analysis. The model investigates the association between
survival time and one or more predictors. The Cox PH model is based on
the hazard function, which is defined as the probability that an
individual will experience the event at a given time, provided that the
event has not occurred yet [@Walters2012]. The outcome variable is the
survival time, which is measured from a defined time until the
occurrence of an event, such as death, or the end of the study period
[@Walters2012]. For each subject, there is time to event (T), which is
the duration from a defined starting point to the event occurrence, or a
censoring time (C), which is the duration until the subject drops out of
the study or the study ends.

The Cox Proportional Hazards (PH) model relies on two main
assumptions: 1) random censoring, and 2) the proportional hazard
assumption [@Patil2019-uk]. The random censoring assumption is satisfied
when patients censored in the study are a random sample from the study
population. There is no statistical test for this assumption, but it can
be ensured through rigorous experimental design. The proportional hazard
assumption requires that the hazard function (hazard ratio) for two
groups (e.g., experimental arm and standard arm) remains proportional,
meaning the hazard ratio is constant over time. This assumption is
violated if a covariate's impact on the outcome changes over the
follow-up period, a common occurrence in biochemical research. When the
proportional hazard assumption is violated, various methods can address
this, including stratifying the model based on variables that violate
the assumption. This stratification allows the baseline hazard to differ
across strata [@Backmann2024-gv]. Another method involves including
interactions between covariates and time, enabling hazard ratios to vary
over time [@Zhang2018-fl]. While baseline hazards are allowed to differ,
the effects of other covariates are presumed consistent across strata. A
limitation of this approach is that the effect of the stratification
variable cannot be tested, as it is not included as a covariate in the
model. Other statistical methods to accommodate non-proportional hazards
are detailed and discussed elsewhere [@Therneau2010-if].

One of the properties of the Cox PH model contributing to its popularity
is that the baseline hazard, i.e., $h_0$, is an unspecified field, which
makes it a semi-parametric model. It is robust and can provide reliable
estimates without specifying the baseline hazard function. Even though
the baseline hazard is not known, it is possible to estimate the
coefficients in the exponential part of the equation. Hence, the effect
of variables included in the model can be estimated [@Kleinbaum1996-do].
Another feature of the Cox PH model is its interpretability. The effect
of covariates on the hazard is represented by hazard ratios, i.e., the
relative likelihood of an event happening in the experimental group with
respect to the standard group.

The Cox PH model is widely used across various fields, including medical
research, epidemiology, business, engineering, and social sciences
[@Kleinbaum1996-do]. It is commonly used to determine survival rates
among cancer patients with different subtypes of cancer
[@Intrieri2023-sp], or between those treated with different treatments
[@Wu2022-gv]. The Cox PH model has also been used effectively in
non-medical research, for example, in the prediction of time to loan
defaults [@Ptak-Chmielewska2024-uf], customer time to churn
[@Wong2011-lf], product lifespan and failure time analysis in
engineering [@Rodriguez-Borbon2017-yz], and for the identification of
factors that influence the time it takes to achieve certain rates of
success, such as vaccination rates, in public health [@Cheng2024-dn].

The current analysis evaluated data from one of the first successful
trials of adjuvant chemotherapy for colon cancer. The colon cancer
patients in the study had their primary treatment of surgery and the
objective was to test whether treatment with either Levamisole or
Levamisole in combination with Fluorouracil (5-FU) chemotherapeutic
agents improves survival. Levamisole is a low-toxicity compound
previously used to treat worm infestations in animals; 5-FU is a
moderately toxic chemotherapy agent. There are two records per person,
one for recurrence and one for death. The purpose of this project is to
compare survival between the untreated (Obs) group vs. those treated
with amisole (Lev), or amisole + 5-FU.

The current analysis evaluated modified data from one of the first
successful trials of adjuvant chemotherapy for colon cancer. The study
involved colon cancer patients who had undergone primary surgical
treatment, and the objective was to test whether treatment with either
Levamisole alone or Levamisole combined with Fluorouracil (5-FU)
chemotherapeutic agents improves survival. Levamisole is a low-toxicity
compound previously used to treat worm infestations in animals; 5-FU is
a moderately toxic chemotherapy agent. There are two records per person,
one for recurrence and one for death. The purpose of this project is to
compare survival between the untreated (Obs) group vs. those treated
with amisole (Lev), or amisole + 5-FU.

## II. Methods

The Cox proportional hazards model was used to model the relationship
between survival time and different colon cancer treatments. In
particular, the survival time between the untreated group (observation)
and those treated with amisole (Lev) or amisole + 5-FU was compared. The
Cox regression model was chosen for this study because it is well-suited
for studying the association between survival time of patients and
predictors, and it allows estimating the risk or hazard of death
increased or decreased due to each treatment relative to no treatment.
The time (in days) until the event, i.e., death, was modeled as a
function of treatment and other variables, including age, sex, and
various tumor characteristics. Significant predictors were included in
the final model.

### IIa. Data Source

The colon cancer cancer dataset is a built-in dataset in the Survival R
package [@survival-package] [@survival-book]. It is closely aligned with
to the original study published in 1991 [@Moertel1990-pd]. The dataset
includes 929 subjects with stage B or C colon cancer, who were
randomized into three treatment groups: Observation, Levamisole (Lev),
or Levamisole + 5-FU. While the dataset contains survival and recurrence
information for these individuals, the current analysis focused on time
to death, so rows representing recurrence time were filtered out. Time
to death is given in days. The median follow-up period was 1,976 days
(5.4 years), with a range of 23-3,329 days. The dataset includes various
patient characteristics, such as demographics and tumor details, as well
as the duration from surgery to trial registration, categorized as
either short or long.

**Table 1: Description of variables included in the dataset**

|  |  |
|:-----------------------|:-----------------------------------------------|
| id: | id |
| study: | 1 for all patients |
| rx: | Treatment - Obs(ervation), Lev(amisole), Lev(amisole)+5-FU |
| sex: | 1=male |
| age: | in years |
| obstruct: | obstruction of colon by tumour |
| perfor: | perforation of colon |
| adhere: | adherence to nearby organs |
| nodes: | number of lymph nodes with detectable cancer |
| time: | days until event or censoring |
| status: | censoring status |
| differ: | differentiation of tumour (1=well, 2=moderate, 3=poor) |
| extent: | Extent of local spread (1=submucosa, 2=muscle, 3=serosa, 4=contiguous structures) |
| surg: | time from surgery to registration (0=short, 1=long) |
| node4: | more than 4 positive lymph nodes |
| etype: | event type: 1=recurrence,2=death |

### IIb. Exploratory data analysis

The data was evaluated for missing values, duplicate entry, and
outliers. The Survminer [@survminer] package was used to plot
Kaplan-Meier curves to visualize survival probability over time for each
of the categorical variables. Multicolinearity was tested using Variant
Inflation Factor (VIF) calculated using MASS package [@MASS].

### IIc. Statistical analysis

The R statistical software version 4.3.2 [@R_citation] was used for all
analysis. The Survival package was used to construct the Cox regression
model [@survival-package] [@survival-book].

Cox regression model is based on the hazard function $h_x(t)$ with
covariates at time t given by:

$h_x(t)=h_0(t)\exp(\beta_1x_1 +\beta_2x_2 + \dots + \beta_p x_p)$

Where:

$h_x(t)$ is the hazard function

$h_0(t)$ is the baseline hazard function

$\beta_1x_1 + \beta_2x_2 + \dots +\beta_p x_p$ represent the linear
combination of covariates and their coefficient

The hazards for the observation vs. amisole (Lev), or amisole + 5-FU
group with covariate values x~1~ and x~2~ are given by:
$hx_1(t)=h_0(t)\exp(\beta_1x_1)$ and $hx_2(t)=h_0(t)\exp(\beta_2x_2)$,
respectively

The hazard ratio is expressed as: HR = $hx_2(t)$ / $hx_1(t)$ =
$\exp[\beta(x_2-x_1)]$

The R MASS package was used for Step-wise variable selection, using
"both" forward and backward variable selection [@MASS]. For Step-wise
selection, stepAIC() function uses AIC (Akaike Information Criterion) as
the measure to add or remove predictors from the model.

### IId. Model evaluation

#### Diagnostic for proportional hazard assumption

The Schoenfeld residual plot was constructed to test Cox proportional
hazards assumption. When the proportional hazards assumption was not met
for any of the covariates, stratification approach was used.

#### K-fold cross-validation

Model performance was evaluated using 5-fold cross-validation using the
boot package [@boot1] [@boot2].

#### Model selection

The Concordance (c-index), AIC, and Baysian Information Criteriod (BIC)
of corresponding models were compared to select the best fit model. The
model with the lowest AIC and BIC values and highest concordance was
selected as the final model.

## III. Analysis and Results

```{r}
#| echo: false
#| message: false
#| warning: false

# Load library
library(dplyr)
library(survminer)
library(survival)
library(janitor)
library(magrittr)
library(car)
library(ggplot2)
library(tidyverse)
library(broom)
library(MASS)
library(boot)
library(gtsummary)
library(knitr)
library(kableExtra)
library(gridExtra)
library(rmarkdown)
library(patchwork)
library(scales)
library(gt)
library(kableExtra)
library(boot)  # for bootstrapping
library(survcomp)  # to calculate c-index
library(caret)


#print(citation("survival"), bibtex=TRUE)

```

### IIIa. Data cleaning and feature engineering

```{r}
#| echo: false
#| message: false
#| code_folding: true
#######################
#Load data
#######################
colon <- as_tibble(colon)
#head(colon)

#filter data to death as the event type. This will create a data table with one row per individual.
colon_surv <- colon%>%filter(etype == 2) 

#######################
#Check missing values
#######################
na_counts <- sapply(colon_surv, function(x)sum(is.na(x)))


# replace NAs with mode
#table(colon_surv$differ)
#mode(colon_surv$differ)
#median(colon_surv$nodes, na.rm= TRUE)

colon_surv$differ <- if_else(is.na(colon_surv$differ), 2,colon_surv$differ) 
colon_surv$nodes <- if_else(is.na(colon_surv$nodes), 2,colon_surv$nodes)

#######################
# Feature engineering
######################
# rename categorical columns to make them descriptive
colon_surv <- colon_surv%>%mutate(differentiation = case_when(differ == 1 ~ "well",
                                                              differ == 2 ~ "moderate",
                                                              differ == 3 ~ "poor"),
                                  local_spread = case_when(extent == 1 ~ "submucosa",
                                                           extent == 2 ~ "muscle",
                                                           extent == 3 ~ "serosa",
                                                           extent == 4 ~ "contiguous"),
                                  surg_to_reg_time = case_when(surg == 0~ "short",
                                                               surg == 1 ~ "long"))





```

The dataset contained missing values in the number of nodes (nodes) and
differentiation (differ) columns. Missing values in the differentiation
and nodes columns were replaced with the mode and median, respectively.
Categories representing different groups in the differentiation, local
spread, and time to surgery (surg) columns were replaced with
descriptive names. Evaluation of the nodes and node4 variables showed
that samples with more than four lymph nodes (node4) had fewer than five
counts in the nodes column, indicating inconsistency between the two
columns. Therefore, the nodes column was not used for further analysis.

### IIIb. Exploratory data analysis

```{r}
#| echo: false
#| message: false
#| include: false

##############################
# Continuous variables
#############################
# age
hist(colon_surv$age)
hist(colon_surv$nodes)
hist(colon_surv$time)

# Evaluate nodes column to investigate outliers

t <- colon_surv%>%filter(node4 ==1) # samples with more than 4 positive lymph nodes
hist(t$nodes) 


###########################
# Categorical variables 
##########################

summary_table <- colon_surv%>%summarise(count =n(),
                                        male = sum(sex), 
                                                       median_age = median(age),
                                                       ct_perforation = sum(perfor),
                                                       ct_adherence_nerby_organ = sum(adhere), censored = sum(status))

summary_table

# frequency tables for categorical variables

# Tumor differentiation

colon_surv %>% 
  tabyl(differentiation, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()

# extent of local spread
colon_surv %>% 
  tabyl(local_spread, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()

# colum obstruction
colon_surv %>% 
  tabyl(obstruct, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()


# colon perforation
colon_surv %>% 
  tabyl(perfor, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()

# Adherance to nearby organs
colon_surv %>% 
  tabyl(adhere, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()

# extent of local tumor spread
colon_surv %>% 
  tabyl(local_spread, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()

# More than 4 lymph nodes with cancer
colon_surv %>% 
  tabyl(node4, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns()


# time from surgery to registration
colon_surv %>% 
  tabyl(surg, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()


########################################################
# Summary statistics grouped by treatment
#######################################################

summary_table <- colon_surv%>%group_by(rx)%>%summarise(count =n(),
                                                       male = sum(sex),
                                                       median_age = median(age),
                                                       ct_perforation = sum(perfor),
                                                       ct_adherence_nerby_organ = sum(adhere),
                                                       perc_male = (male/count)*100,
                                                       iqr_age = IQR(age))
# summary_table


```

```{r}
#| echo: false
#| message: false
#| include: true
############################
# Data visualization
############################

# Subset data for modeling
df <- colon_surv %>% dplyr::select(!c(id, study, etype, differ, extent, surg_to_reg_time, nodes))

df <- df %>%
  mutate(time_years = time / 365,
         time_years_round = round(time_years, 0))

# Define categorical and continuous columns
categorical_cols <- c("rx", "sex", "obstruct", "perfor", "adhere", "status", "surg", "node4", "differentiation", "local_spread")
continuous_cols <- c("age", "time_years")

# Function to create percentage bar plots for categorical variables
plot_categorical <- function(col) {
  ggplot(df, aes(x = as.factor(.data[[col]]))) +
    geom_bar(aes(y = after_stat(count) / sum(after_stat(count)))) +
    geom_text(aes(label = scales::percent(after_stat(count) / sum(after_stat(count)), accuracy = 1), 
                  y = after_stat(count) / sum(after_stat(count))), 
              stat = "count", vjust = -0.5, size = 3) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.1))) +
    labs(y = "Percentage", x = NULL) +
    ggtitle(paste("", col)) +
    theme(plot.margin = unit(c(0.001, 0.001, 0.001, 0.001), "cm"),
          plot.title = element_text(size = 14),
          axis.title.y = element_text(size = 12),
          axis.text.x = element_text(angle = 45, hjust = 1)) 
}

# Function to create histogram plots for continuous variables
plot_continuous <- function(col) {
  ggplot(df, aes(x = .data[[col]])) +
    geom_histogram(bins = 30, fill = "blue", color = "black") +
    scale_x_continuous(labels = scales::number_format(accuracy = 1)) +
    scale_y_continuous(labels = scales::number_format(accuracy = 1)) +
    ggtitle(paste("Distribution of", col, "stratified by status")) +
    theme(plot.margin = unit(c(0.001, 0.001, 0.001, 0.01), "cm"),
          plot.title = element_text(size = 14),
          axis.title.y = element_text(size = 12),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    facet_wrap(~status)
}

# Create individual plots
plot_list <- lapply(categorical_cols, plot_categorical)
continuous_plot_list <- lapply(continuous_cols, plot_continuous)

# Combine all plots into a single grid
all_plots <- wrap_plots(plot_list, ncol = 5)
cont_plot <- wrap_plots(continuous_plot_list, ncol = 2)

# Increase plot area
options(repr.plot.width = 20, repr.plot.height = 15)

# Display the plots
all_plots
cont_plot

```

**Figure 1: Exploratory data analysis**<br>

Exploratory data visualization (Figure 1) showed that the study
participants were equally randomized into three treatment groups, with
approximately 33% of individuals in the observation, Levamisole (Lev),
or Lev + 5-FU groups. Fifty-two percent of the participants were male.
The dataset is balanced with respect to the outcome (status), with 51%
censored and 49% having died before the end of the follow-up period.

Regarding tumor characteristics, obstruction of the colon by the tumor
(obstruct), perforation of the colon (perfor), and tumor adherence to
nearby organs (adhere) were observed in 19%, 3%, and 15% of the
patients, respectively. About 27% of the participants had more than four
tumor-positive lymph nodes (node 4) and a long time from surgery to
registration (surg). Overall, most categories of tumor characteristics
are well represented, with about 15% of the participants having these
characteristics, except for perforation, differentiation, and local
spread.

Age is normally distributed. Evaluation of time to death or censoring
showed a bimodal distribution. Stratifying the time variable by outcome
(status) revealed that the majority of the patients died within three
years of the study period.

```{=html}
<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-bobw{font-weight:bold;text-align:center;vertical-align:bottom}
.tg .tg-7zrl{text-align:left;vertical-align:bottom}
.tg .tg-8d8j{text-align:center;vertical-align:bottom}
.tg .tg-j6zm{font-weight:bold;text-align:left;vertical-align:bottom}
</style>
```

**Table 2: Patient characteristics**<br>

|   |   | Observation (%) | Amisole (%) | Amisole + 5-FU (%) |
|---------------|---------------|:-------------:|:-------------:|:-------------:|
|  |  | N=315 | N=310 | N=304 |
| **Demographics** |  |  |  |  |
|  | Male | 166 (52.3) | 177 (57.1) | 141 |
|  | Median age (years) \[IQR\] | 60 \[53,68\] | 61 \[53,69\] | 61 \[52,70\] |
| **Cancer characteristics** |  |  |  |  |
|  | Colon obstruction | 63 (20.0) | 63 (20.3) | 54 (17.8) |
|  | Colon perforation | 9 (2.9) | 10 (3.2) | 8 (2.6) |
|  | Adherence to nearby organs | 47 (14.9) | 49 (15.8) | 39 (12.8) |
| Differentiation of tumor |  |  |  |  |
|  | Well | 27 (8.6) | 37 (11.9) | 29 (9.5) |
|  | Moderate | 236 (74.9) | 229 (73.9) | 221 (72.7) |
|  | Poor | 52 (16.5) | 44 (14.2) | 54 (17.8) |
| Extent of local spread |  |  |  |  |
|  | Contiguous | 20 (6.3) | 12 (3.9) | 11 (3.6) |
|  | Muscle | 38 (12.1) | 36 (11.6) | 32 (10.5) |
|  | Serosa | 249 (79.0) | 259 (83.5) | 251 (82.6) |
|  | Submucosa | 8 (2.5) | 3 (1.0) | 10 (3.3) |
| More than 4 lymph nodes with cancer | Yes | 87 (27.6) | 89 (28.7) | 79 (26.0) |
| Short time from surgery to registration (%) | Yes | 91 (28.9) | 80 (25.8) | 76 (25.0) |

As shown in Table 2, the baseline patient characteristics were very
similar between the three treatment groups, suggesting that differences
in survival time between the groups are not likely attributed to
differences at the start of the study.

### IIIc. Survival curves

```{r}
#| echo: false
#| message: false
#| warning: false
#| include: true

colon_surv$idcount <- c(1:length(colon_surv$id))

# Order by survival time and create an order variable:
colon_surv <- colon_surv[order(-colon_surv$time, colon_surv$status),]
colon_surv$order <- c(1:length(colon_surv$idcount))

# Create a data frame for vline intercepts
vline_data <- data.frame(
  status = unique(colon_surv$status),
  vline = c(2331, 775)  # Example values for each status
)

# Create a data frame for text labels
label_data <- data.frame(
  status = unique(colon_surv$status),
  x = 2600,  # X position for the label
  y = 750,   # Y position for the label
  label = "Median Survival Time"
)

ggplot(data=colon_surv, aes(x=time, y=order)) +
  geom_rect(xmin=23, xmax=colon_surv$time, ymin=colon_surv$order, ymax=colon_surv$order+1, colour="lightblue") +
  geom_rect(xmin=colon_surv$time-2, xmax=colon_surv$time, ymin=colon_surv$order, ymax=colon_surv$order+1, fill=factor(colon_surv$status+1)) +
  geom_vline(data=vline_data, aes(xintercept=vline), linetype="solid") +  # Add vertical lines with different values
  geom_text(data=label_data, aes(x=x, y=y, label=label)) +  # Add text labels
  scale_x_continuous(breaks=seq(20, 3330, 650)) +
  scale_y_continuous(breaks=seq(0, max(colon_surv$order), 100)) +  # Add y-axis labels in increments of 50
  xlab("Survival Time (Days)") + 
  ylab("Participants (ordered by survival time)") +
  ggtitle("Survival Times for Participants Who Were Censored or Died") +
  facet_wrap(~status) +
  theme_classic() +
  theme(
    legend.position="none",
    panel.grid.major=element_blank(),
    panel.grid.minor=element_blank(),
    panel.background=element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black")
  )
```

**Figure 2: Censoring (left) or survival (right) time of patients**<br>

The median survival time of patients who were censored (left) or died
(right) was 2,331 and 775 days, respectively. These results show that
patients who were not censored died early during the follow-up period.

```{r}
#| echo: false
#| message: false
#| warning: false
#| include: false
#| 
t <- df%>%filter(status == 0)
t2 <- df%>%filter(status == 1)
median(t$time)
median(t2$time)
```

**Stratified survival curves**

```{r}

#| echo: false
#| message: false
#| warning: false
#| include: true


f1 <- survfit(Surv(time, status) ~ adhere, data = df)
f2 <- survfit(Surv(time, status) ~ rx, data = df)
f3 <- survfit(Surv(time, status) ~ sex, data = df)
f4 <- survfit(Surv(time, status) ~ obstruct, data = df)
f5 <- survfit(Surv(time, status) ~ perfor, data = df)
f6 <- survfit(Surv(time, status) ~ node4, data = df)
f7 <- survfit(Surv(time, status) ~ differentiation, data = df)
f8 <- survfit(Surv(time, status) ~ local_spread, data = df)
f9 <- survfit(Surv(time, status) ~ surg, data = df)

fits <- list(adhere = f1, rx = f2, sex =f3, obstruct = f4 , perfor=f5 , node4 =f6 , differentiation =f7 , local_spread =f8 ,  surg= f9)

# Visualize

legend.title <- list("adhere", "rx", "sex", "obstruct", "perfor", "node4", "differentiation", "local_spread", "surg")
ggsurvplot_list(fits, df,  pval=TRUE) #legend.title = legend.title,

```

**Figure 3: Survival curves stratified by categorical variables**<br>

There is a statistically significant difference in survival times
between two or more groups being compared, as shown in the survival
curves stratified by adherence to nearby organs, rx, obstruction,
presence of four or more tumor-positive nodes, differentiation, local
spread, and surgery. Survival time is not significantly different
between males and females or among those with or without tumor
perforation. The vertical lines on the Kaplan-Meier curve, which show
censoring data points, are enriched after about 1,800 days, suggesting
that many participants were lost to follow-up after this period.

Patient characteristics that showed significant differences in survival
time between two categories are important predictors of survival time.
However, the Kaplan-Meier curve does not adjust for other variables;
therefore, some of the predictors may not be significant when included
in the multivariate Cox regression model.

### IIId. Cox regression models

```{r}
#| echo: false
#| message: false
#| warning: false
#| include: false
# Subset data for modeling
df <- colon_surv%>%dplyr::select(!c(id,study,etype,differ, extent,surg_to_reg_time, idcount, order, nodes))

```

#### Model_1: Base Model

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: true

Model_1 <- coxph(Surv(time, status) ~ 1, data = df)
c_index_model_1 <- concordance(Model_1)
cat("Concordance of the base model:",c_index_model_1$concordance)

```

#### Model_2: Univariate analysis

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: true

# Univariate analysis
Model_2 <- coxph(Surv(time, status) ~ rx, data = df)


# Create the regression table and add concordance statistic
summary_table <- tbl_regression(Model_2, exponentiate = TRUE) %>%
  add_glance_source_note(
    label = list(concordance = "Concordance"),
    include = c("concordance")
  ) %>%
  modify_table_styling(
    columns = p.value,
    rows = p.value < 0.05,
    text_format = "bold"
  )

# Convert to gt table, increase font size, and adjust width
gt_table <- as_gt(summary_table) %>%
  gt::tab_options(
    table.font.size = px(17),  # Increase font size
    table.align = "center",      # Align table to the left
    table.width = pct(50)     # Make the table wider
  ) %>%
  gt::tab_header(
    title = gt::md("**Table 3. Model_2: Univariate model**")
  )

# Print the gt table
gt_table

```

The coefficient of Lev is not significant, suggesting that there is no
evidence that this treatment affects survival time compared to
observation. however Lev+5Fu is significant (p=0.00175), indicating that
the treatment Lev +5Fu has a statistically significant effect on
survival time compared to the reference group. The negative sign
indicates that this treatment group has a lower hazard and likely a
longer survival time. The hazard ratio for Lex+5FU (0.690), indicating
the risk of death is about 31% lower compared to the observation group.

#### Model_2: Cox proportional hazard assumption test

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: true
#| 

zph_test <- cox.zph(Model_2)

# Convert the Schoenfeld residuals test results to a data frame
zph_df <- as.data.frame(zph_test$table)
zph_df$Variable <- rownames(zph_df)

zph_df <- zph_df %>%
  mutate(
    chisq = round(chisq, 3),
    p = round(p, 3)
  )

zph_df %>%
  kbl(caption = "Table 4: Schoenfeld Residuals Test Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, align = "center")

# plot the Schoenfeld residuals
plot(zph_test)

```

**Figure 4: Schoenfeld residal plot of rx variable included in
Model_2**<br>

The Schoenfeld residal plot shows that the residuals are scattered
randomly and the smooth trend line is horizontal near 0. This suggests
that the hazard ratio for rx (treatment status) is constant over time
and the proportional hazard assumption is met. The global p-value is
\>0.05, indicating that the the assumption is met.

#### Model_3: All predictors

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: true
#| 
# Full variables: All predictors
Model_3<- coxph(Surv(time, status) ~ rx+ age + sex + perfor + adhere + surg + obstruct + differentiation + node4+ local_spread, data = df)

summary_table <- tbl_regression(Model_3, exponentiate = TRUE) %>%
  add_glance_source_note(
    label = list(concordance = "Concordance"),
    include = c("concordance")
  ) %>%
  modify_table_styling(
    columns = p.value,
    rows = p.value < 0.05,
    text_format = "bold"
  )

gt_table <- as_gt(summary_table) %>%
  tab_options(
    table.font.size = px(17),  # Reduce font size
    table.align = "center",    # Align table to the center
    table.width = pct(50),     # Make the table wider
    data_row.padding = px(2)   # Reduce row padding
  ) %>%
  gt::tab_header(
    title = gt::md("**Table 5. Model_3: All predictors**")
  )

gt_table

```

The summary of Model_3 shows that when all variables are included in the
model, age, sex, perforation, and adherence are not significant
predictors. In contrast, rx, surg, obstruct, differentiation, node4, and
local spread are significant predictors. The concordance of the
multivariable model (0.674) is higher than that of the univariate model
(concordance = 0.53), suggesting that the multivariate model is a better
fit. However, the model concordance did not improve when removing
predictors that were not significant in Model_3.

#### Stepwise variable selection:

A step-wise variable selection approach was used to confirm that the
significant predictors in Model_3 are the best set of predictors. The
stepAIC() function from the MASS package was used for stepwise variable
selection, utilizing the Akaike Information Criterion (AIC) to add or
remove predictors from the model.

```{R}

# Model_2 includes all variables
Model_2 <- coxph(Surv(time, status) ~ rx+ age + sex + perfor + adhere + surg + obstruct + differentiation + node4+
              local_spread, data = df)  

# stepwise selection
stepwise_model <- stepAIC(Model_2, direction = "both")

# Extract the coefficients of the selected model
selected_variables <- coef(stepwise_model)

# Print the selected variables
print(selected_variables)
```

Based on a step-wise selection, rx, age, surg, obstruct,
differentiation, node4 and local spread were selected for inclusion in
the final model. Age is a significant predictor in the step-wise
selected variables, but not in the full variable model (Model_2).

#### Model_4: Stepwise-Selected variables

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: true
Model_4 <- coxph(Surv(time, status) ~ rx + age + surg + obstruct + 
    differentiation + node4 + local_spread, data = df)

summary_table <- tbl_regression(Model_4, exponentiate = TRUE) %>%
  add_glance_source_note(
    label = list(concordance = "Concordance"),
    include = c("concordance")
  ) %>%
  modify_table_styling(
    columns = p.value,
    rows = p.value < 0.05,
    text_format = "bold"
  )

gt_table <- as_gt(summary_table) %>%
  tab_options(
    table.font.size = px(17),  # Reduce font size
    table.align = "center",    # Align table to the center
    table.width = pct(50),     # Make the table wider
    data_row.padding = px(2)   # Reduce row padding
  ) %>%
  gt::tab_header(
    title = gt::md("**Table 6. Model_4: Stepwise-selected variables**")
  )

gt_table
```

The concordance of Model_4 did not improve compared to Model_3, which
included all predictors. However, Model_4 is less complex as it includes
a smaller number of variables (see model comparison table below).

#### Model_4: Cox proportional hazard assumption test

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: true

 # final model with step wise variable selection
zph_test <- cox.zph(Model_4)

# Convert the Schoenfeld residuals test results to a data frame

zph_df <- as.data.frame(zph_test$table)
zph_df$Variable <- rownames(zph_df)

zph_df <- zph_df %>%
  mutate(
    chisq = round(chisq, 3),
    p = round(p, 3)
  )

zph_df %>%
  kbl(caption = "Table 7. Schoenfeld Residuals Test for Model_4") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, align = "center")

```

Model_4 does not meet the Cox proportional hazards assumption, as
indicated by p-values less than 0.05 in the Schoenfeld residuals test.
The differentiation, node4, and obstruct variables did not meet the
proportional hazards assumption, suggesting that the effects of these
variables are not constant over time. To address this, a stratification
approach will be used, where the model will be stratified by the
variables that did not meet the proportional hazards assumption.

#### Model_5: Stratified Model

```{r}
#| echo: True
#| message: false
#| warning: false
#| include: true
#| 
Model_5 <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 +
              local_spread, data = df)

summary_table <- tbl_regression(Model_5, exponentiate = TRUE) %>%
  add_glance_source_note(
    label = list(concordance = "Concordance"),
    include = c("concordance")
  ) %>%
  modify_table_styling(
    columns = p.value,
    rows = p.value < 0.05,
    text_format = "bold"
  )

gt_table <- as_gt(summary_table) %>%
  tab_options(
    table.font.size = px(17),  # Reduce font size
    table.align = "center",    # Align table to the center
    table.width = pct(50),     # Make the table wider
    data_row.padding = px(2)   # Reduce row padding
  )%>%
  gt::tab_header(
    title = gt::md("**Table 8. Model_5: Stratified Model**")
  )
gt_table
```

#### Model_5 Cox proportional hazard assumption test

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: true

zph_test <- cox.zph(Model_5)
# Convert the Schoenfeld residuals test results to a data frame
zph_df <- as.data.frame(zph_test$table)
zph_df$Variable <- rownames(zph_df)
zph_df <- zph_df %>%
  mutate(
    chisq = round(chisq, 3),
    p = round(p, 3)
  )

zph_df %>%
  kbl(caption = "Table 9. Schoenfeld Residuals Test for Model_5") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, align = "center")

# plot the Schoenfeld residuals
plot(zph_test)
```

**Figure 5: Schoenfeld residal plot of variables included in
Model_5**<br>

After model stratification by obstruct and differentiation, the
proportional hazards assumption is met, as the global p-value is greater
than 0.05. Node4 slightly violates the assumption, but the final model
is not stratified by node4 because the model concordance is attenuated
when stratifying by node4.

#### Evaluate multicollinearity of variables in Model_5

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: true

vif <- vif(Model_5)
print(vif)
```

None of the variables has VIF \>5, therefore multicollinearity is not a
problem in Model_5

#### Model comparison

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: true

# Fit the models and store them in a list
models <- list(Model_1, Model_2, Model_3, Model_4, Model_5)

# Add descriptions for each model
descriptions <- c(
  "Base model",
  "Treatment",
  "Full variables",
  "Stepwise-selected variables",
  "Stratified"
)


# Create a data frame to store results
results <- data.frame(
  Model = character(),
  Description = character(),
  AIC = numeric(),
  BIC = numeric(),
  C_Index = numeric(),
  stringsAsFactors = FALSE
)

# Function to calculate and store metrics for each model
for (i in seq_along(models)) {
  model <- models[[i]]
  
  # Extract AIC and BIC
  aic <- AIC(model)
  bic <- BIC(model)
  
  # Add C-index
  c_index <- concordance(model)$concordance
  
  # Append results to the data frame
  results <- rbind(results, data.frame(
    Model = paste("Model", i),
    Description = descriptions[i],
    AIC = aic,
    BIC = bic,
    C_Index = round(c_index, 3)
  ))
}

# Print the table using kable and kableExtra
results %>%
  kbl(caption = "Table 10. Model Evaluation Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:5, width = "10em") %>%
  kable_styling(position = "center")
```

Based on the model evaluation metrics, Model_5 has the smallest AIC and
BIC values and achieved the highest concordance (c-index). Although
Model_2 and Model_5 have the same concordance, Model_5 is the best model
as indicated by the smaller AIC and BIC values compared to Model_2.
Five-fold cross-validation was performed to test the robustness of
Model_5.

#### K-fold cross validation

```{r}
#| echo: true
#| message: false
#| warning: false
#| include: true

set.seed(1234)

# Cox model 
cox_model <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 +
              local_spread, data = df)


# calculate the original c-index
c_index_original <- concordance(cox_model)
cat("original c-index:", c_index_original$concordance, "\n")

# create a function for calculating c-index in each fold using concordance()
cox_cindex <- function(train_data, test_data) {
  fit <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 + local_spread, data = train_data)
  # Calculate concordance on test data
  c_index <- concordance(fit, newdata = test_data)$concordance
  
  return(c_index)
}

# perform 5-fold cross-validation with stratification
K <- 5
folds <- createFolds(c(df$status, df$differentiation, df$rx), k = K, list = TRUE, returnTrain = TRUE)
cv_c_indices <- sapply(folds, function(train_indices) {
  train_data <- df[train_indices, ]
  test_data <- df[-train_indices, ]
  cox_cindex(train_data, test_data) # use the concordance function inside cox_cindex
})

# Print cross-validated c-indices
print(cv_c_indices)

# cross-validation c-indices
cat("cross-validated c-Indices for each fold:", cv_c_indices, "\n")
cat("mean cross-validated c-Index:", mean(cv_c_indices), "\n")

# plot cross-validation c-indices
plot(cv_c_indices, type = "b", xlab = "Fold", ylab = "c-index", main = "c-index across folds")

```

**Figure 6: Concordance across the 5 folds**<br>

Model_5's c-index (0.674) and mean cross-validation c-index (0.675) are
very similar, suggesting that the final stratified model is stable and
not overfitting.

## IV. Conclusions

-   Treatment with Levamisole + 5-FU decreases the hazard of death from
    colon cancer by 29.5% (p=0.003).

-   Having more than 4 tumor positive lymph nodes significantly
    increases the hazard of death by 150% (p\<0.0001).

-   Compared to patients with tumors that have spread to contiguous
    organs, those with tumors localized in the submucosa, muscle, and
    serosa have significantly reduced hazard rates by 76% (p = 0.007),
    66% (p \< 0.001), and 43% (p = 0.004), respectively.

-   Having long wait from surgery to trial registration increases the
    hazard rate by about 29.5% (p=0.012), suggesting that starting
    chemotherapy treatment within a short period of time after surgery
    improves survival.

-   For each additional year of age, the hazard rate increases by about
    0.87% (p = 0.034), suggesting that age has minimal effect on
    survival.

-   The concordance of the model (0.674) indicates moderate predictive
    accuracy for survival time.

-   Other variables that were not included in the study may contribute
    to survival time.
