---
title: "Cox regression modeling of survival after chemotherapy for colon cancer"
author: "Hermela Shimelis"
date: today
format:
  html:
    code-fold: False
    css: custom.css
    toc: true
    toc_float:
      collapsed: TRUE
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
csl: ieee.csl
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72    
---

## Data: Survival after chemotherapy for Stage B/C colon cancer @survival-book @survival-package

## Introduction

Cox regression is one of the most widely used statistical model to study survival analysis. In survival analysis, the outcome variable is the
survival time, T, measured from a defined time, such as time of
treatment initiation, or time of disease diagnosis to the occurrence of
an event, such as death [@Walters2012].

Sure, here's a summary of why linear regression and logistic regression cannot be used for survival analysis:

Outcome Variable: In survival analysis, the outcome is the survival time (T), measured from a specific starting point (e.g., diagnosis, treatment start) to an event (often death).
Quantitative Nature: The survival time is a quantitative variable, which might suggest the use of standard linear regression models.
Survival Status: If the focus is on whether a subject is alive or dead at a specific time (τ), one might consider logistic regression for the binary outcome.
Right-Censoring: Survival analysis often involves right-censoring, where the exact survival time is unknown for some subjects because they are still alive or have left the study.
Impact of Censoring: Right-censoring means that standard linear or logistic regression methods are not directly applicable.
Independent Censoring Assumption: Valid inference requires the assumption that censoring is independent, meaning the fact that a subject is still alive at time t should not provide additional information about their survival time beyond T > t.
Untestable Assumption: This assumption is crucial but cannot be tested with the available incomplete data.
Need for Alternative Methods: Due to right-censoring and the need for the independent censoring assumption, alternative methods like the Cox proportional hazards model are necessary for survival analysis.

### Description

These are data from one of the first successful trials of adjuvant
chemotherapy for colon cancer. Levamisole is a low-toxicity compound
previously used to treat worm infestations in animals; 5-FU is a
moderately toxic (as these things go) chemotherapy agent. There are two
records per person, one for recurrence and one for death.

The purpose of this project is to compare survival between the untreated
(Obs) group vs those treated with amisole (Lev), or amisole + 5-FU.

Column names:

|           |                                                                                   |
|:----------------|:------------------------------------------------------|
| id:       | id                                                                                |
| study:    | 1 for all patients                                                                |
| rx:       | Treatment - Obs(ervation), Lev(amisole), Lev(amisole)+5-FU                        |
| sex:      | 1=male                                                                            |
| age:      | in years                                                                          |
| obstruct: | obstruction of colon by tumour                                                    |
| perfor:   | perforation of colon                                                              |
| adhere:   | adherence to nearby organs                                                        |
| nodes:    | number of lymph nodes with detectable cancer                                      |
| time:     | days until event or censoring                                                     |
| status:   | censoring status                                                                  |
| differ:   | differentiation of tumour (1=well, 2=moderate, 3=poor)                            |
| extent:   | Extent of local spread (1=submucosa, 2=muscle, 3=serosa, 4=contiguous structures) |
| surg:     | time from surgery to registration (0=short, 1=long)                               |
| node4:    | more than 4 positive lymph nodes                                                  |
| etype:    | event type: 1=recurrence,2=death                                                  |
|           |                                                                                   |

```{r}
#| echo: true
#| message: false
#| warning: false



# Load library
library(dplyr)
library(survival)
library(janitor)
library(magrittr)
library(car)
library(ggplot2)
library(tidyverse)
library(broom)
library(MASS)
library(boot)
library(gtsummary)
library(knitr)
library(kableExtra)
library(gridExtra)

#print(citation("survival"), bibtex=TRUE)

```

```{r}
#| echo: true
#| message: false

#Load data
colon <- as_tibble(colon)
head(colon)
```

Since the current analysis is focused on survival, filter data to death
as the event type. This will create a data table with one row per
individual.

```{r}
colon_surv <- colon%>%filter(etype == 2) 

```

## I. Exploratory data analysis

**Check missing values**

```{r}

na_counts <- sapply(colon_surv, function(x)sum(is.na(x)))
na_counts

# replace NAs with mode
table(colon_surv$differ)
mode(colon_surv$differ)
median(colon_surv$nodes, na.rm= TRUE)

colon_surv$differ <- if_else(is.na(colon_surv$differ), 2,colon_surv$differ) 
colon_surv$nodes <- if_else(is.na(colon_surv$nodes), 2,colon_surv$nodes)
```

**Insight:** only nodes and differ columns have NA values. Replacing the
23 NAs in differ column with mode, and replace NAs in nodes with median.

**Evaluate continuous variables**

```{r}
#| echo: true
#| message: false

# age
hist(colon_surv$age)
hist(colon_surv$nodes)
hist(colon_surv$time)


```

**Insight:** Age is normally distributed. Number of nodes is skewed to
the right. Time is fairly normally distributed with most the individuals
having event time between 500-3000 days.

Evaluate nodes column to investigate outliers

```{r}
t <- colon_surv%>%filter(node4 ==1) # samples with more than 4 positive lymph nodes
hist(t$nodes) 


```

**Insight:** samples with greater than 4 lymph nodes have less than 5
count in nodes column, so the two columns are not consistent. Therefore,
nodes column will not be used for further analysis.

**Evaluate categorical variables**

```{r }
#| echo: true
#| message: false
 
summary_table <- colon_surv%>%summarise(count =n(),
                                        male = sum(sex), 
                                                       median_age = median(age),
                                                       ct_perforation = sum(perfor),
                                                       ct_adherence_nerby_organ = sum(adhere), censored = sum(status))

summary_table
```

**Insight:** Total number of participants: 929. About half of the
participants are male and about half were censored, while the other half
died.

```{r}
# rename categorical columns to make them descriptive
colon_surv <- colon_surv%>%mutate(differentiation = case_when(differ == 1 ~ "well",
                                                              differ == 2 ~ "moderate",
                                                              differ == 3 ~ "poor"),
                                  local_spread = case_when(extent == 1 ~ "submucosa",
                                                           extent == 2 ~ "muscle",
                                                           extent == 3 ~ "serosa",
                                                           extent == 4 ~ "contiguous"),
                                  surg_to_reg_time = case_when(surg == 0~ "short",
                                                               surg == 1 ~ "long"))



```

**Frequency tables for categorical variables**

```{r}
#| warning: false

# frequency tables for categorical variables
# Tumor differentiation

colon_surv %>% 
  tabyl(differentiation, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()

# extent of local spread
colon_surv %>% 
  tabyl(local_spread, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()

# colum obstruction
colon_surv %>% 
  tabyl(obstruct, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()


# colon perforation
colon_surv %>% 
  tabyl(perfor, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()

# Adherance to nearby organs
colon_surv %>% 
  tabyl(adhere, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()

# extent of local tumor spread
colon_surv %>% 
  tabyl(local_spread, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()

# More than 4 lymph nodes with cancer
colon_surv %>% 
  tabyl(node4, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns()


# time from surgery to registration
colon_surv %>% 
  tabyl(surg, rx) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns()
```

**Summary statistics grouped by treatment**

```{r #| echo: true#| message: false}

summary_table <- colon_surv%>%group_by(rx)%>%summarise(count =n(),
                                                       male = sum(sex),
                                                       median_age = median(age),
                                                       ct_perforation = sum(perfor),
                                                       ct_adherence_nerby_organ = sum(adhere),
                                                       perc_male = (male/count)*100,
                                                       iqr_age = IQR(age))
summary_table


```

**Insight:** Each treatment group had about 300 participants. Median
age, number of participants with perforation and adherence are similar
between the three groups.

## II. Table 1: Description of the study population

```{=html}
<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-bobw{font-weight:bold;text-align:center;vertical-align:bottom}
.tg .tg-7zrl{text-align:left;vertical-align:bottom}
.tg .tg-8d8j{text-align:center;vertical-align:bottom}
.tg .tg-j6zm{font-weight:bold;text-align:left;vertical-align:bottom}
</style>
```
|                                             |                            | Observation (%) | Amisole (%)  | Amisole + 5-FU (%) |
|-----------------|--------------|:------------:|:------------:|:------------:|
|                                             |                            |      N=315      |    N=310     |       N=304        |
| **Demographics**                            |                            |                 |              |                    |
|                                             | Male                       |   166 (52.3)    |  177 (57.1)  |        141         |
|                                             | Median age (years) \[IQR\] |  60 \[53,68\]   | 61 \[53,69\] |    61 \[52,70\]    |
| **Cancer characteristics**                  |                            |                 |              |                    |
|                                             | Colon obstruction          |    63 (20.0)    |  63 (20.3)   |     54 (17.8)      |
|                                             | Colon perforation          |     9 (2.9)     |   10 (3.2)   |      8 (2.6)       |
|                                             | Adherence to nearby organs |    47 (14.9)    |  49 (15.8)   |     39 (12.8)      |
| Differentiation of tumor                    |                            |                 |              |                    |
|                                             | Well                       |    27 (8.6)     |  37 (11.9)   |      29 (9.5)      |
|                                             | Moderate                   |   236 (74.9)    |  229 (73.9)  |     221 (72.7)     |
|                                             | Poor                       |    52 (16.5)    |  44 (14.2)   |     54 (17.8)      |
| Extent of local spread                      |                            |                 |              |                    |
|                                             | Contiguous                 |    20 (6.3)     |   12 (3.9)   |      11 (3.6)      |
|                                             | Muscle                     |    38 (12.1)    |  36 (11.6)   |     32 (10.5)      |
|                                             | Serosa                     |   249 (79.0)    |  259 (83.5)  |     251 (82.6)     |
|                                             | Submucosa                  |     8 (2.5)     |   3 (1.0)    |      10 (3.3)      |
| More than 4 lymph nodes with cancer         | Yes                        |    87 (27.6)    |  89 (28.7)   |     79 (26.0)      |
| Short time from surgery to registration (%) | Yes                        |    91 (28.9)    |  80 (25.8)   |     76 (25.0)      |

# III. Methods

The Cox proportional hazards model was used to model the relationship
between survival time and different lung cancer treatments. In
particlular the survival time will be compared between the untreated
group (observation) vs. those treated with amisole (Lev), or amisole +
5-FU. The Cox regression model was chosen for this study because it is
useful for studying association between survival time of patients and
predictors and allows estimating the relative risk or hazard ratios due
to the covariates, i.e., treatment status. The time (in days) until
event, i.e, death, will be modeled as a function of treatment and other
variables, including age, sex, and various tumor characteristics.
Significant predictors were included in the final model.

## Statistical analysis

The R statistical software version 4.3.2 [@R_citation] was used for all
analysis. The Survival package was used to construct the Cox regression
model [@survival-package] [@survival-book].

Cox regression model is based on the hazard function $h_x(t)$ with
covariates at time t given by:

$h_x(t)=h_0(t)\exp(\beta_1x_1 +\beta_2x_2 + \dots + \beta_p x_p)$

Where:

$h_x(t)$ is the hazard function

$h_0(t)$ is the baseline hazard function

$\beta_1x_1 + \beta_2x_2 + \dots +\beta_p x_p$ represent the linear
combination of covariates and their coefficient

The hazards for the observation vs. amisole (Lev), or amisole + 5-FU
group with covariate values x~1~ and x~2~ are given by:
$hx_1(t)=h_0(t)\exp(\beta_1x_1)$ and $hx_2(t)=h_0(t)\exp(\beta_2x_2)$,
respectively

The hazard ratio is expressed as: HR = $hx_2(t)$ / $hx_1(t)$ =
$\exp[\beta(x_2-x_1)]$

The Schoenfeld residual plot was constructed to test Cox proportional
hazards assumption. When the proportional hazards assumpiton was not met
for any of the covariates, stratification approach was explored. The
Survminer [@survminer] package was used to plot the Kaplan-Meier curve
to visualize the survival probability over time for each treatment
group.

Multicolinearity was tested using Variant Inflation Factor (VIF)
calculated using MASS package [@MASS].

The R MASS package was used for Stepwise model selection, using "both"
forward and backward variable selection [@MASS]. For Stepwise selection,
stepAIC() function uses AIC (Akaike Information Criterion) as the
measure to add or remove predictors from the model. Model performance
was evaluated using 100-fold cross-validation using the boot package
[@boot1] [@boot2].

# IV. Analysis: Cox regression model

### Survival curve

```{r}
#| echo: true
#| message: false
#| warning: false
# Create new incremental count id
colon_surv$idcount <- c(1:length(colon_surv$id))

# Order by survival time and create an order variable:
colon_surv <- colon_surv[order(-colon_surv$time, colon_surv$status),]
colon_surv$order <- c(1:length(colon_surv$idcount))

ggplot(data=colon_surv, aes(x=time, y=order)) +
geom_rect(xmin=23,xmax=colon_surv$time,ymin=colon_surv$order,ymax=colon_surv$order+1, colour="lightgray") +
geom_rect(xmin=colon_surv$time-2,xmax=colon_surv$time,ymin=colon_surv$order,ymax=colon_surv$order+1,
          fill=factor(colon_surv$status+1)) +
geom_vline(xintercept= 1976,linetype="solid") +
scale_x_continuous(breaks=seq(20,3330,650)) +
geom_text(aes(2600, 750, label="Median Survival Time")) +
xlab("Survival Time (Days)") + ylab("Participants (ordered by survival time)") +
ggtitle("Survival Times for Participant") +
theme_classic() +
theme(legend.position="none",
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      panel.background=element_blank(),
      axis.line.x = element_line(color = "black"),
      axis.line.y = element_line(color = "black"))

```

### Survival curve stratified by treatment group

```{r}
#| echo: true
#| message: false
#| warning: false

library(survminer)
library(survival)

# Estimate the median survival time among the three groups
survfit(Surv(time,status) ~ rx, data = colon_surv)

# count the number of events after 2080 days, which is the median survival time among the observation group
tt <- colon_surv%>%filter(time > 2083)%>% group_by(rx)%>%summarise(ct = n(),
                                                                   death = sum(status))
# Plot survival curve
fit <- survfit(Surv(time,status) ~ rx, data = colon_surv)
ggsurvplot(fit, data=colon_surv, risk.table = TRUE)

# Estimate the probability of surviving beyond 3000 days
summary(survfit(Surv(time, status) ~ rx, data = colon_surv), times = 3000)

# compare significant differences in survival times between the three groups
survdiff(Surv(time, status)~ rx, data = colon_surv)


```

**Insight:** Based on the survival curve, the mediant survival time for
the observation group is 2083 days. However, the median survival of Lev
and Lev+5Fu group cannot be estimated, because there are too few events
after 2083 days, which is the median survival time in the observation
group.

The time for 50% survival probability of the group treated with Lev+5Fu
is over 3000 days while the survival time for the observation and Lev
group is around 2080 days. The probability of surviving to 3000 days
among the Lev+5FU group is 56% (95% CI: 50-63), compared to 41% among
the observation group.

The survival time is significantly different (P=0.003) between the three
groups.

**Insight:** None of the variables have VIF values above 5, therefore
there is no multicollinearity

## Cox regression models

```{r}
# Subset data for modeling
df <- colon_surv%>%dplyr::select(!c(id,study,etype,differ, extent,surg_to_reg_time, idcount, order, nodes))


```

### Base Model

```{r}
m0 <- coxph(Surv(time, status) ~ 1, data = df)
summary_m0 = summary(m0)

c_index_m0 <- concordance(m0)

cat("Concordance of the base model:",c_index_m0$concordance)

# Calculate the baseline hazard function
baseline_hazard <- basehaz(m0, centered = FALSE)

# Print the baseline hazard
print(baseline_hazard)

# Plot the cumulative baseline hazard function
ggplot(baseline_hazard, aes(x = time, y = hazard)) +
  geom_step() +
  labs(title = "Cumulative Baseline Hazard Function",
       x = "Time",
       y = "Cumulative Baseline Hazard") +
  theme_minimal()
```

### Univariate analysis

```{r}

#| echo: true
#| message: false
#| warning: false

# Univariate analysis
m1 <- coxph(Surv(time, status) ~ rx, data = df)
summary(m1)

c_index_m1 <- concordance(m1)
cat("Concordance of the univariate model:",c_index_m1$concordance)

anova(m0, m1) # Addition of rx variable significantly improved base model
```

**Insight:** the coefficient of Lev is not significant, suggesting that
there is no evidence that this treatment affects survival time compared
to observation. however Lev+5Fu is significant (p=0.00175), indicating
that the treatment Lev +5Fu has a statistically significant effect on
survival time compared to the reference group. The negative sign
indicates that this treatment group has a lower hazard and likely a
longer survival time.

The hazard ratio for Lex+5FU (0.690), indicating the risk of death is
about 31% lower compared to the observation group.

The p-values indicate that the model is significant.

### Test the Cox proportional hazard assumption of m1

```{r}

cox.zph(m1)
zph_test <- cox.zph(m1)

print(zph_test)

# plot the Schoenfeld residuals
plot(zph_test)

```

**Insight:** The Schoenfeld residal plot shows that the residuals are
scattered randomly and the smooth trend line is horizontal near 0. This
suggests that the hazard ratio for rx (treatment status) is constant
over time and the proportional hazard assumption is met. The global
p-value is \>0.05, indicating that the the assumption is met.

### Multivariate analysis

```{r}
#| echo: true
#| message: false
#| warning: false
#| 
# Include all variables to determine which predictors are significant.

names(df)
# multivariate analysis
m2 <- coxph(Surv(time, status) ~ rx+ age + sex + perfor + adhere + surg + obstruct + differentiation + node4+
              local_spread, data = df)  
summary(m2)

c_index_m2 <- concordance(m2)
cat("Concordance of the multivariate model:",c_index_m2$concordance)

# Determine significant predictors
anova(m2)

anova(m0, m1, m2)

```

**Insight:** When all variables are included in the model, the anova
test indicates that rx, adhere, surg, obstruct, differentiation, node4
and local spread are significant predictors. Additionally, model
concordance did not improve when removing predictors that were not
significant in m2

The concordance of the multivariable model, 0.674, is higher than the
univariate model (m1, concordance =0.53), suggesting that the
multivariate model is a better fit model.

### Evaluate multicollinearity using Variance Inflation Factor (VIF)

```{r}

vif <- vif(m2)
print(vif)
```

#### Evaluate significance of predictors. Model survival while including different cancer characteristics as predictors separately to identify significance predictors.

```{r}
# model including all variables
m2 <- coxph(Surv(time, status) ~ rx+ age + sex + perfor + adhere + surg + obstruct + differentiation + node4+
              local_spread, data = df)  


# Treatment
m2a <- coxph(Surv(time, status) ~ rx, data = df) # significant
summary(m2a)

# Demographics
m2b <- coxph(Surv(time, status) ~ age + sex, data = df) # not significant
summary(m2b)


# cancer characteristics
m2c <- coxph(Surv(time, status) ~ perfor + adhere + obstruct, data = df) # adhere and obstruct are significant
summary(m2c)

# Differentiation of tumor
m2d <- coxph(Surv(time, status) ~ differentiation, data = df) # significant
summary(m2d)

# Extent of local spread
m2e <- coxph(Surv(time, status) ~ local_spread, data = df) # significant
summary(m2e)


# short time from surgery to registration
m2f <- coxph(Surv(time, status) ~ surg, data = df) # significant
summary(m2f)


# include predictors significant in the model which included all predictors (m2)
m3 <- coxph(Surv(time, status) ~ rx + surg + obstruct + differentiation + node4
              + local_spread, data = df)

summary(m3)
c_index_m3 <- concordance(m3)
cat("Concordance of the multivariate model2:",c_index_m3$concordance)


anova(m0, m1, m2, m2a, m2b, m2c, m2d, m2e, m2f, m3)
```

**Insight:** rx, adhere, surg, obstruct, differentiation and
local_spread are significant predictors. However, the model concordance
is low (\~0.5) when each was included separately. Model m2, which
included all predictors, followed by model 3 (including selected
significant predictors) have the highest concordance.

### Perform Stepwise variable selection:

```{R}

library(MASS)       # for stepwise regression
#### Use the MASS package stepAIC() function for stepwise selection by using AIC (Akaike Information Criterion) as the measure to add or remove predictors from the model.

# model including all variables
m2 <- coxph(Surv(time, status) ~ rx+ age + sex + perfor + adhere + surg + obstruct + differentiation + node4+
              local_spread, data = df)  

# stepwise selection
stepwise_model <- stepAIC(m2, direction = "both")

summary(stepwise_model)

summary_table <- tbl_regression(m2 , exponentiate = TRUE)

# Print the summary table
summary_table

# Extract the coefficients of the selected model
selected_variables <- coef(stepwise_model)

# Print the selected variables
print(selected_variables)
```

```{r}

# Multivariate model including variables selected based on stepwise variable selection. The same variables were significant based on anova test of the model that included all variables.

m4 <- coxph(Surv(time, status) ~ rx + age + surg + obstruct + 
    differentiation + node4 + local_spread, data = df)
summary(m4)
anova(m4)

cox_summary <- tidy(m4)

cox_summary

c_index_m4 <- concordance(m4)
cat("Concordance of the model with multivariate stepwise v_select:",c_index_m4$concordance)

```

### Test whether Proportional hazard assumptions are met for model 4 predictors

```{R}

cox.zph(m4) # final model with stepwise variable selection
zph_test <- cox.zph(m4)

print(zph_test)

# plot the Schoenfeld residuals
plot(zph_test)


```

**Insight:** Differentiation, node4 and obstruct variables did not meet
proportional hazards assumption.

### Stratify model by variables violating roportional hazard assumption

```{r}
m5 <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 +
              local_spread, data = df)

summary(m5)

cox.zph(m5) # final model with stratification by variables violating proportional hazard assumption
zph_test <- cox.zph(m5)

print(zph_test)

# plot the Schoenfeld residuals
plot(zph_test)
```

**Insight:** After model stratification by obstruct and differentiation,
the proportional hazard assumption is met as the global p \>0.05. Node4
slightly violates assumption, but the final model is not stratified by
node 4 because the model concordance is attenuated when stratifying by
node4.

## Model comparision

```{r}
library(survival)
library(dplyr)
library(knitr)
library(kableExtra)

# Fit the models and store them in a list
models <- list(m0, m1, m2, m3, m4)

# Add descriptions for each model
descriptions <- c(
  "Model 0 - Base model",
  "Model 1 - Treatment",
  "Model 2 - Full variables",
  "Model 3 - Selected stepwisely",
  "Model 4 - model 4 Stratified"
)


# Create a data frame to store results
results <- data.frame(
  Model = character(),
  Description = character(),
  AIC = numeric(),
  BIC = numeric(),
  C_Index = numeric(),
  stringsAsFactors = FALSE
)

# Function to calculate and store metrics for each model
for (i in seq_along(models)) {
  model <- models[[i]]
  
  # Extract AIC and BIC
  aic <- AIC(model)
  bic <- BIC(model)
  
  # Add C-index
  c_index <- concordance(model)$concordance
  
  # Append results to the data frame
  results <- rbind(results, data.frame(
    Model = paste("Model", i),
    Description = descriptions[i],
    AIC = aic,
    BIC = bic,
    C_Index = round(c_index, 3)
  ))
}

# Print the table using kable and kableExtra
results %>%
  kbl(caption = "Model Evaluation Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:5, width = "10em") %>%
  kable_styling(position = "center")
```

## K-fold cross validation

```{r}
library(survival)
library(boot)  # for bootstrapping
library(survcomp)  # to calculate c-index
library(caret)


set.seed(1234)

# Cox model 
cox_model <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 +
              local_spread, data = df)


# calculate the original c-index
c_index_original <- concordance(cox_model)
cat("original c-index:", c_index_original$concordance, "\n")

# create a function for calculating c-index in each fold using concordance()
cox_cindex <- function(train_data, test_data) {
  fit <- coxph(Surv(time, status) ~ rx + age + surg + strata(obstruct) + strata(differentiation) + node4 + local_spread, data = train_data)
  # Calculate concordance on test data
  c_index <- concordance(fit, newdata = test_data)$concordance
  
  return(c_index)
}

# perform 5-fold cross-validation with stratification
K <- 5
folds <- createFolds(c(df$status, df$differentiation, df$rx), k = K, list = TRUE, returnTrain = TRUE)
cv_c_indices <- sapply(folds, function(train_indices) {
  train_data <- df[train_indices, ]
  test_data <- df[-train_indices, ]
  cox_cindex(train_data, test_data) # use the concordance function inside cox_cindex
})

# Print cross-validated c-indices
print(cv_c_indices)

# cross-validation c-indices
cat("cross-validated c-Indices for each fold:", cv_c_indices, "\n")
cat("mean cross-validated c-Index:", mean(cv_c_indices), "\n")

# plot cross-validation c-indices
plot(cv_c_indices, type = "b", xlab = "Fold", ylab = "c-index", main = "c-index across folds")

```

**Insight:** The original model c-index (0.674) and mean
cross-validation c-index (0.675) is very similar, suggesting the the
final stratified model is stable and is not overfitting.

# V. Results

### Table 2. Univariate model: Survival after Chemotherapy for stage B/C Colon Cancer

| Treatment      | Coefficient | Hazard ratio | 95% CI_upper | 95% CI_lower | P-value |
|------------|:----------:|:----------:|:----------:|:----------:|:----------:|
| Amisole (Lev)  |   -0.027    |    0.974     |    0.784     |    1.209     |  0.809  |
| Amisole + 5-FU |   -0.372    |    0.690     |    0.546     |    0.870     |  0.002  |

### Table 3. Multivariate model: Survival after Chemotherapy for stage B/C Colon Cancer

| Treatment              | Coefficient | Hazard ratio | 95% CI_upper | 95% CI_lower |    P-value     |
|------------|:----------:|:----------:|:----------:|:----------:|:----------:|
| Amisole (Lev)          |   -0.011    |    0.989     |    0.795     |    1.231     |     0.923      |
| Amisole + 5-FU         |   -0.376    |    0.687     |    0.543     |    0.868     |     0.002      |
| Age                    |    0.007    |    1.007     |    0.999     |    1.015     |     0.069      |
| Surge                  |    0.244    |    1.276     |    1.042     |    1.562     |     0.018      |
| Obstruction of colon   |    0.283    |    1.327     |    1.057     |    1.667     |     0.015      |
| Differentiat ion_poor  |    0.374    |    1.453     |    1.145     |    1.844     |     0.002      |
| Differentiat ion_well  |    0.069    |    1.072     |    0.774     |    1.483     |     0.677      |
| More than 4 nodes (+)  |    0.930    |    2.534     |    2.089     |    3.074     | 3.75 x 10^-21^ |
| Local spread_muscle    |   -0.996    |    0.370     |    0.225     |    0.606     | 7.85 x 10^-5^  |
| Local spread_serosa    |   -0.501    |    0.606     |    0.414     |    0.886     |     0.010      |
| Local spread_submucosa |   -1.322    |    0.267     |    0.093     |    0.763     |     0.014      |
